\documentclass[11pt]{article}

\usepackage{fullpage}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mdwlist}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}

\input{macros}




\begin{document}

\ptitle{Problem Set 1}{January 22, 2015}


This problem set is due on Friday, January 30, by 5pm. Please submit your solution online using bcourses,
as a pdf file. 

You can type your solution, or handwrite it. If you handwrite it, then either
scan it or take a good resolution picture of each page and then collate the pictures
and export them to a {\em single} pdf file.

\bigskip

\hrule

\section*{Problem 1: Repeat (15/100)}


\noindent Let $\Sigma=\{a,b\}$. Let $L$ be a language over $\Sigma$,
consisting of strings of length at least 2,
where the first letter is the same as the last letter,
and the second letter is the same as the second to last letter.
For example, $ a \not\in L$, $b \not\in L$, $aa \in L$, $aaa \in L$, $aba \in L$, $bbaabba \not\in L$.\\

\noindent Design a DFA that accepts $L$.


\section*{Problem 2: Multiples of 3 (25/100)}

\noindent Let $\Sigma = \{0, 1\}$, and $L$ be the language of 
all binary strings that are the binary representations of an integer multiple of 3 (leading zeroes are
allowed).  For example, $0 \in L$, $101 \not \in L$, $1100 \in L$.\\
 
\noindent Design a DFA that accepts $L$


\section*{Problem 3: Prefix, Suffix, Mid (60/100)}
Let $\Sigma$ be an alphabet. {\color{Gray}Ex: $\Sigma=\{a, b, ..., z\}$}.\\
Let $\circ$ be the concatenation operator. {\color{Gray}Ex: ``hello'' $\circ$ ``world'' = ``helloworld''}\\
\noindent Let $L$ be a language over alphabet $\Sigma$, where
prefix(L) = $\left\{x \mid \exists z:\, x \circ z \in L\right\}$,
suffix(L) = $\left\{z \mid \exists x:\, x \circ z \in L\right\}$, and
mid(L) = $\left\{y \mid \exists x, z:\, x \circ y \circ z \in L\right\}$.

\begin{itemize}
\item Given a DFA that accepts $L$, design a DFA or NFA that accepts prefix$(L)$. (20 points)
\item Given a DFA that accepts $L$, design a DFA or NFA that accepts suffix$(L)$. (20 points)
\item Given a DFA that accepts $L$, design a DFA or NFA that accepts mid$(L)$. (20 points)
\end{itemize}


\end{document}


%\section*{Problem 2: Union, Intersection, Complement, Difference}
%Let $L$ and $R$ be regualr languages over alphabet $\Sigma$.
%\begin{enumerate}
%\item Prove that $L \cup R = \{ x | x \in L \lor x \in R\}$ is regular. (1 point)
%\item Prove that $L \cap R = \{ x | x \in L \land x \in R\}$ is regular. (1 point)
%\item Prove that $\overline{L} = \{ x | x \in \Sigma^*, x \not\in L\}$ is regular. (1 point)
%\item Prove that $L \setminus R = \{ x | x \in L, x \not\in R\}$ is regular. (1 point)
%\end{enumerate}
%
%For proving statement $i$, you are allowed to use statements
%$1, ..., i-1$, even if you did not prove them.
