\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mdwlist}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{enumerate}


\input{macros}


\begin{document}
\ptitle{Problem Set 9}{April 10, 2015}
This problem set is due on Friday, April 17, by 5pm. Please submit your solution online using bcourses,
as a pdf file.
You can type your solution, or handwrite it. If you handwrite it, then either
scan it or take a good resolution picture of each page and then collate the pictures
and export them to a {\em single} pdf file.
\\

\noindent When asked to prove that a problem is $NL$-complete,
make sure to include a proof that the problem is in $NL$. When
specifying  an $NL$ algorithm, preferably give pseudo-code
for it, and use terminology like ``guess $x\in S$" to specify a
non-deterministic step in which variable $x$ can take any of the
values in the set $S$. 
\bigskip
\hrule



\section*{Problem 1 (35/100)}
Consider the {\em acceptance problem for
NFA}, that is the language 

\[ A_{NFA} := \{ (M,w) : M \mbox{ is an NFA } \wedge M \mbox{ accepts } w \} \]

\noindent Show that $A_{NFA}$ is $NL$-complete under $\leq_m^{\log}$ reductions.

\noindent [You can use the fact that the reachability problem for directed
graphs is $NL$-complete]

\section*{Problem 2 (35/100)}
Consider the DAG-Reach problem defined as follows

\[ DAG-Reach := \{ (G,s,t) : G \mbox{ G is a DAG} \wedge t \mbox{ is
reachable from } s \} \]

\noindent (where DAG stands for directed acyclic graph). Show that
DAG-Reach is $NL$-complete. (This shows that the reachability
problem in directed graphs remains $NL$-complete even when
restricted to acyclic graphs.)

\noindent [Hint: create multiple copies of each vertex/configuration, and 
arrange them in layers.]

\section*{Problem 3 (30/100)}
Consider the language

\[ StrongConn := \{ G : G \mbox { is a strongly connected directed graph} \} \]

\noindent Recall that a graph $G=(V,E)$ is strongly connected if for every two
vertices $u,v$ there is a part from $u$ to $v$ and also a path
from $v$ to $u$. Prove that StrongConn is $NL$-complete.

\noindent [Hint: reduce from DAG-Reach]
\end{document}